\documentclass[12pt]{article}

\usepackage{amsmath, amssymb, xypic}
\usepackage[margin=1in]{geometry}

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

\begin{document}
\begin{center}
Assignment 1\\
Intro to Modern Analysis
\end{center}

\noindent \textbf{1.} For a subset $A$ of a set $X$, let $A^c$ denote the complement of $A$ given by $A^c = X \setminus A$. Let $I$ be an index set, and let $\{A_i\}_{i \in I}$ be a collection of subsets of $X$ indexed by $I$. Prove or disprove.  
\begin{enumerate}
\item[(a)] We have 
\[
\left(\bigcup_{i \in I}A_i\right)^c = \bigcap_{i \in I}A_i^c.
\]
\item[(b)] We have 
\[
\left(\bigcap_{i \in I}A_i\right)^c = \bigcup_{i \in I}A_i^c.
\]
\end{enumerate}

\medskip

\noindent \textbf{2.} Let $\{A_n\}_{n \in \mathbb{N}}$ be a sequence of nonempty subsets of $\mathbb{R}$ which are ``nested'' in the sense that 
\[
A_0 \supset A_1 \supset A_2 \supset \cdots.
\]
Is the intersection 
\[
\bigcap_{n \in\mathbb{N}} A_n
\]
nonempty? Prove or disprove. 

\medskip

\noindent \textbf{3.} Let $A$ denote the subset of $\mathbb{R}$ given by  
\[
A = \left\{\frac{|\cos(n)|}{n+1} : n \in \mathbb{N} \right\}.
\]
Find with proof the values $\sup A$ and $\inf A$, whenever either exists. 

\medskip


\noindent \textbf{4.} Let $A$ be a nonempty subset of $\mathbb{R}$ bounded from above, and let $x_0$ be an upper bound of $A$.  Show that $x_0$ is equal to $\sup A$ if and only if for each $\epsilon > 0$ there is an element $x \in A$ such that $x_0 - x < \epsilon$. 

\medskip 

\noindent \textbf{5.} Let $A$ be a nonempty set of \emph{integers} that is bounded from above. Show that $A$ has a largest element. (Hint: Let $x_0 = \sup A$. First show that $x_0$ is an integer by showing that if not, then there are integers $m,n \in A$ satisfying $x_0 - 1 < m < n < x_0$, which is an impossible statement [why?]. Then show that $x_0$ belongs to $A$.)


\medskip




\noindent \textbf{6.} Let $A, B$ be nonempty subsets of $\mathbb{R}$ each bounded from above. Prove or disprove. 
\begin{enumerate}
\item[(a)] $\sup (A \cup B) = \sup\{ \sup A, \sup B \}$.
\item[(b)] $\sup(A \cap B) = \inf\{ \sup A, \sup B \}$. 
\end{enumerate}

\medskip

\noindent \textbf{7.} For a real number $t$, let $|t|$ denote the absolute value of $t$ defined by  
\[
|t| = \begin{cases}
t & t \geqslant 0 \\
-t & t \leqslant 0
\end{cases}.
\] 
\begin{enumerate}
\item[(a)] For a real number $t$ and a nonnegative number $a$, show that $|t|\leqslant a$ if and only if $-a \leqslant t \leqslant a$. 
\item[(b)] Let $x$ and $y$ be points of $\mathbb{R}^n$. Show that 
\[
\big\lvert\lVert{x}\rVert - \lVert{y}\rVert\big\rvert \leqslant \norm{x -y}. 
\]
\end{enumerate}



\end{document}