\documentclass[12pt]{article}

\usepackage{amsmath, amssymb, xypic}
\usepackage[margin=1in]{geometry}

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

\begin{document}
\begin{center}
Assignment 2\\
Intro to Modern Analysis
\end{center}


\noindent \textbf{1.} Let $A$ be a set, and let $P(A)$ be the collection of all subsets of $A$. Is there a bijection from $A$ onto $P(A)$? 

\medskip

\noindent \textbf{2.} Let $M$ be an infinite set, and let $A$ be a countable set. Is there a bijection from $M$ onto $M \cup A$? 

\medskip 

\noindent \textbf{3.} Let $\{A_k\}_{k \geqslant 1}$ be a sequence of subsets of a metric space. Prove or disprove. 
\begin{enumerate}
\item[(a)] For each integer $N > 0$, we have 
\[
\bigcup_{k=1}^N \overline{A_k} \subset \overline{\bigcup_{k=1}^N A_k} 
\]
\item[(b)] For each integer $N > 0$, we have 
\[
\bigcup_{k=1}^N \overline{A_k} = \overline{\bigcup_{k=1}^N A_k}.
\]
\item[(c)] We have 
\[
\bigcup_{k=1}^\infty \overline{A_k} \subset \overline{\bigcup_{k=1}^\infty A_k}.
\]
\item[(d)] We have 
\[
\bigcup_{k=1}^\infty \overline{A_k} = \overline{\bigcup_{k=1}^\infty A_k} 
\]
\end{enumerate}

\medskip

\noindent \textbf{4.} Let $A^\circ$ denote the set of interior points of $A$. 
\begin{enumerate}
\item[(a)] Prove that $A^\circ$ is open. 
\item[(b)] Prove that $A$ is open if and only if $A = A^\circ$. 
\item[(c)] If $B \subset A$ and $B$ is open, prove that $B \subset A^\circ$. 
\item[(d)] Prove that the complement of $A^\circ$ is the closure of the complement of $A$. 
\end{enumerate}

\medskip

\noindent \textbf{5.} Let $X$ be the interval $[0,2) \subset \mathbb{R}$. The restriction of the usual metric on $\mathbb{R}$ to $X$ is a metric on $X$. 
\begin{enumerate}
\item[(a)] Is the set $[0,1)$ open relative to $X$? 
\item[(b)] Is the set $[1,2)$ closed relative to $X$?
\item[(c)] Is the set $[1,2)$ compact relative to $X$?
\end{enumerate}

\medskip

\noindent \textbf{6.} Give an example of an open cover of $(0,1)$ which admits no finite subcover. 

\medskip

\noindent \textbf{7.} Let $f(x) = x^2$. 
\begin{enumerate}
\item[(a)] Let $x_n = f(2 + \tfrac{1}{n})$. Does $x_n$ converge? Prove your answer is correct. 
\item[(b)] Let $y_n = f(n)$. Does $y_n$ converge? Prove your answer is correct. 
\end{enumerate}

\medskip


\noindent \textbf{8.} Let $\{A_n\}_{n \geqslant 1}$ be a sequence of open dense subsets of $\mathbb{R}$. Let $A$ denote the intersection 
\[
A = \cap_{n \geqslant 1} A_n.
\]
\begin{enumerate}
\item[(a)] Let $U$ be an open subset of $\mathbb{R}$. Show that there is a sequence of points $x_n \in U \cap A_k$ together with a sequence of radii $0 < r_n < 1/n$ such that $\overline{B_{r_{n+1}}(x_{n+1})} \subset B_{r_n}(x_n) \cap A_n$.
\item[(b)] Show that the sequence $x_n$ is Cauchy, and hence converges to a point $x$ of $\mathbb{R}$. 
\item[(c)] Show that $x \in U \cap A$. 
\item[(d)] Conclude that $A$ is dense in $\mathbb{R}$. 
\end{enumerate}


\noindent \textbf{9.} For $x = (x_1, \ldots, x_n) \in \mathbb{R}^n$ and $p > 0$, write 
\[
\norm{x}_p = \left(\sum_{i=1}^n |x_i|^p\right)^{1/p}.
\]
Fix a point $x = (x_1, \ldots, x_n) \in \mathbb{R}^n$, and for each positive integer $k > 0$, let $a_k$ denote the sequence of real numbers $a_k = \norm{x}_{k}$.  Show that 
\[
\lim_{k \to \infty} a_k = \max_{1 \leqslant i \leqslant n} |x_i|. 
\]

\medskip 

\noindent\textbf{10.} In the notation of the previous question, define a function $d_p : \mathbb{R}^n \times \mathbb{R}^n \to \mathbb{R}$ by the rule 
\[
d_p(x,y) = \norm{x - y}_p.
\]
Is $(\mathbb{R}^n, d_p)$ a metric space for $p \in (0,1)$ and $n > 1$? 



\end{document}