\documentclass{article}

\usepackage{amsmath, amssymb, xypic}
\usepackage[margin=.8in]{geometry}

\begin{document}
\begin{center}
MATH S1202: Calculus IV \\
Summer 2018 \\
Section 001
\end{center}


\noindent \textbf{Instructor:} Mitchell Faulk \\
\textbf{Time:} MTWR 1:00-2:35pm  \\
\textbf{Place:} Math 417 \\
\textbf{Office Hours:} MW 10:10-11:25am in Math 610 \\
\textbf{Teaching Assistant:} Jacob Austin (ja3067@columbia.edu) \\
\textbf{Webpage:} http://math.columbia.edu/$\sim$faulk/CalcIVSummer2018

\section*{Prerequisites}

Calculus III or the equivalent is required. 

\section*{Description and goals}



General goals include 
\begin{itemize}
\item To develop further techniques for solving higher-dimensional problems occurring in the sciences or engineering, especially those problems involving integration
\item To develop further a geometric intuition about vectors and geometric objects in two- or three-dimensional space
\end{itemize}
The main concepts covered include double and triple integrals; change of variables; line and surface integrals; vector fields and vector integral calculus; Green's theorem; divergence theorem; and Stokes' theorem. 




\section*{Policies}

\subsection*{Grading}

The course will be graded on a scale of 450 points, distributed as follows 

\vspace{3mm}


\noindent Homework: 100 points (10 points each)\\
Quizzes: 100 points (20 points each) \\
Midterm: 100 points \\
Final: 150 points



\subsection*{Homework}

Most homework will be assigned via WebAssign, due online before class starts. Homework will be due typically twice a week (on Mondays and Wednesdays), except possibly during those weeks of exams. 

\subsection*{Quizzes}

There will be one quiz per week, at the beginning of class on Thursdays. Each quiz will last about 30 minutes.  

If you receive less than a full score on the quizzes, then you can earn back up to half of the points you missed by submitting a new set of solutions. Your solutions must be submitted within one day of receiving your returned quiz.  

\subsection*{Exams}

There will be one midterm exam and one final exam. The dates are 

\vspace{3mm}

\noindent Midterm exam date: 06/05 \\
Final exam date: 06/28

\vspace{3mm}

\noindent If you have a conflict with either of these dates, you \textbf{must} contact me ahead of time to make arrangements. (At least a week in advance is ideal.) 


\subsection*{Textbook and WebAssign}

You are required to buy WebAssign. The class key is \textbf{columbia 1909 7798}. Use your UNI as your username so that I can identify you. 

On the ``Calculus Classess'' webpage, there is information about purchasing the textbook and WebAssign at a reduced price. 

\subsection*{Students with disabilities}

In order to receive any disability-related accommodations, students must be registered with Disability Services (DS). Students that have, or think they may have, a disability are encouraged to contact DS for more information regarding policies and services available. 

\section*{Syllabus and schedule}

\begin{tabular}{| l | l | l | l |}\hline
Date & Material & Textbook & Announcements \\ \hline\hline
05/21 & Double integrals over rectangles & \S 15.1  &  \\
05/22 & Double integrals over general regions & \S 15.2  &  \\
05/23 & Polar coordinates, applications & \S 15.3, 15.4 & HW1 due \\
05/24 & Surface area, triple integrals & \S 15.5, 15.6 & Quiz 1 \\ \hline \hline  
05/28 & HOLIDAY & HOLIDAY & HOLIDAY \\
05/29 & Cylindrical coordinates, spherical coordinates & \S 15.7, 15.8 &   \\
05/30 & Change of variables & \S 15.8, 15.9 & HW2 due \\
05/31 & Change of variables & \S 15.9 & Quiz 2\\ \hline \hline 
06/04 & Review  &  &  HW3 due  \\
06/05 & \textbf{Midterm exam} & &  \\
06/06 & Vector fields & \S 16.1 & HW4 due \\
06/07 & Line integrals & \S 16.2 & Quiz 3 \\ \hline \hline 
06/11 & Fundamental theorem for line integrals & \S 16.3 & HW5 due \\ 
06/12 & Green's Theorem & \S 16.4 &  \\
06/13 & Curl and divergence & \S 16.5 & HW6 due \\
06/14 & Parametric surfaces, surface area & \S 16.6 & Quiz 4\\ \hline \hline 
06/18 & Surface integrals & \S 16.7 & HW7 due \\
06/19 & Stokes' Theorem & \S 16.8 & \\
06/20 & Divergence theorem & \S 16.9 & HW8 due \\
06/21 & Complex functions &  & Quiz 5\\ \hline \hline 
06/25 & Cauchy-Riemann equations & & HW9 due \\
06/26 & Contour integrals and Cauchy's theorem & & \\
06/27 & Review & & HW10 due \\
06/28 & \textbf{Final exam} & & \\ \hline
\end{tabular}


\section*{Other advice}

This is a fast-paced course, so keeping up to date with material is important. The quizzes are designed to help you in this task. In addition, I strongly encourage you to work on homework assignments early, read the textbook before lecture, and ask questions during lecture if you have any. 


\end{document}